
% Cal Poly Thesis
% 
% based on UC Thesis format
%
% modified by Mark Barry 2/07.
%

\documentclass[12pt]{ucthesis}

\usepackage{ifpdf} 
\newif\ifpdf
\ifx\pdfoutput\undefined
    \pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue \fi

\usepackage{url}
\usepackage{multicol}
\ifpdf

    \usepackage[pdftex]{graphicx}
    % Update title and author below...
    \usepackage[pdftex,plainpages=false,breaklinks=true,colorlinks=true,urlcolor=blue,citecolor=blue,%
                                       linkcolor=blue,bookmarks=true,bookmarksopen=true,%
                                       bookmarksopenlevel=3,pdfstartview=FitV,
                                       pdfauthor={Forrest Reiling},
                                       pdftitle={Extending Windowing Systems to Three Dimensions},
                                       pdfkeywords={thesis, masters, cal poly}
                                       ]{hyperref}
    %Options with pdfstartview are FitV, FitB and FitH
    \pdfcompresslevel=1

\else
    \usepackage{graphicx}
\fi

\usepackage{hyperref}
\hypersetup{
	linktoc=all,
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}



\usepackage{titlesec}
% \titleformat{\chapter}[display]% OLD
%     {\normalfont\huge\bfseries}{\chaptertitlename\ \thechapter}{20pt}{\Huge}% OLD
% \titlespacing*{\chapter}{0pt}{50pt}{40pt}% OLD
\titleformat{\chapter}[display]% NEW
    {\normalfont\centering}{\chaptertitlename\ \thechapter}{12pt}{}% NEW
\titlespacing*{\chapter}{0pt}{30pt}{20pt}% NEW

%\titleformat{\section}[block]{first}{label}{12pt}

\titleformat{\section}{}{\thesection}{1em}{}
\titleformat{\subsection}{}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}{}{\theparagraph}{1em}{}

\usepackage[font={}]{caption}

%\renewcommand{\cftchapleader}{\cftdotfill{\cftdotsep}} % for chapters
%\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}} 


\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[letterpaper]{geometry}
\usepackage[overload]{textcase}
%\usepackage[toc,page]{appendix}

\usepackage{tabularx}
\usepackage{mdframed}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{fixltx2e}


\usepackage{rotating}

\usepackage{enumitem}
\setlist{nolistsep}


\usepackage{float}
\floatstyle{boxed}
%\restylefloat{table}

\bibliographystyle{abbrv}

\setlength{\parindent}{0.25in} \setlength{\parskip}{6pt}

\geometry{verbose,nohead,tmargin=1.25in,bmargin=1in,lmargin=1.5in,rmargin=1.3in}

\setcounter{tocdepth}{4}
\setcounter{secnumdepth}{4}

% Different font in captions (single-spaced, bold) ------------
%\newcommand{\captionfonts}{\small\bf\ssp}
\newcommand{\captionfonts}{}

\makeatletter  % Allow the use of @ in command names
\long\def\@makecaption#1#2{%
  \vskip\abovecaptionskip
  \sbox\@tempboxa{{\captionfonts #1: #2}}%
  \ifdim \wd\@tempboxa >\hsize
    {\captionfonts #1: #2\par}
  \else
    \hbox to\hsize{\hfil\box\@tempboxa\hfil}%
  \fi
  \vskip\belowcaptionskip}
\makeatother   % Cancel the effect of \makeatletter
% ---------------------------------------



\begin{document}

% Declarations for Front Matter

% Update fields below!
\title{Optimizing Lempel-Ziv Factorization for the GPU Architecture}
\author{Bryan Ching}
\degreemonth{June} \degreeyear{2014} \degree{Master of Science}
\defensemonth{June} \defenseyear{2014}
\numberofmembers{3} \chair{Professor Chris Lupo, Ph.D.,\newline Department of Computer Science} \othermemberA{Professor John Seng, Ph.D.,\newline Department of Computer Science} \othermemberB{Professor Zachary N J Peterson, Ph.D.,\newline Department of Computer Science} \field{Computer Science} \campus{San Luis Obispo}
\copyrightyears{seven}



\maketitle

\begin{frontmatter}

% Custom made for Cal Poly (by Mark Barry, modified by Andrew Tsui).
\copyrightpage

% Custom made for Cal Poly (by Andrew Tsui).
\committeemembershippage

\begin{abstract}
Lossless data compression is used to reduce storage requirements, allowing for the relief of I/O channels and better utilization of bandwidth.
The Lempel-Ziv lossless compression algorithms form the basis for many of the most commonly used compression schemes.
General purpose computing on graphic processing units (GPGPUs) allows us to take advantage of the massively parallel nature of GPUs for computations other that their original purpose of rendering graphics. 
Our work targets the use of GPUs for general lossless data compression.
Specifically, we developed and ported an algorithm that constructs the Lempel-Ziv factorization directly on the GPU.
Our implementation bypasses the sequential nature of the LZ factorization and attempts to compute the factorization in parallel.
By breaking down the LZ factorization into what we call the PLZ, we are able to outperform the fastest serial CPU implementations by up to 24x and perform comparatively to a parallel multicore CPU implementation.
To achieve these speeds, our implementation outputted LZ factorizations that were on average only 0.01 percent greater than the optimal solution that what could be computed sequentially.

We are also able to reevaluate the fastest GPU suffix array construction algorithm, which is needed to compute the LZ factorization.
We are able to find speedups of up to 5x over the fastest CPU implementations.

\end{abstract}


\begin{acknowledgements}

Thanks to:

\begin{itemize}
\item My advisor Chris Lupo, for all of his guidance and patience and for introducing me to the GPGPU.
\item My parents, my family, and everyone who has supported me to where I am today.
\end{itemize}

\end{acknowledgements}

\tableofcontents

\listoftables

\listoffigures

\end{frontmatter}

\pagestyle{plain}

\renewcommand{\baselinestretch}{1.66}


% ------------- Main chapters here --------------------
\include{chapters/Introduction}
\include{chapters/Background}
\include{chapters/Implementation}
\include{chapters/Results}
\include{chapters/Conclusion}
\include{chapters/FutureWork}

% ------------- End main chapters ----------------------

\clearpage
\nocite{*}
%\bibliographystyle{plain}
\bibliography{Bibliography}
%\addcontentsline{toc}{chapter}{Bibliography}

%\begin{appendix}
%%\addcontentsline{toc}{chapter}{\appendixnamelower}
%\include{Appendix}
%\end{appendix}

\end{document}
